---
title: "Getting Started with xnatR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with xnatR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

xnatR is an R package for interacting with [XNAT](https://www.xnat.org/) neuroimaging data archives. XNAT is an open-source imaging informatics platform used by research institutions worldwide to manage and share neuroimaging data.

This vignette covers:

- Authentication and credential management
- Navigating the XNAT data hierarchy
- Downloading data
- Advanced searching
- Working with alias tokens

## Installation

```{r installation}
# Install from GitHub
# install.packages("remotes")
remotes::install_github("bbuchsbaum/xnatR")
```

```{r setup}
library(xnatR)
```

## Authentication

Before accessing any XNAT data, you must authenticate with an XNAT server. xnatR supports multiple authentication methods.

### Direct Credentials

The simplest approach is to provide credentials directly:

```{r auth-direct}
authenticate_xnat(
  base_url = "https://central.xnat.org",
  username = "guest",
  password = "guest"
)
```

### Environment Variables

For scripts and automated workflows, use environment variables:
- `XNATR_HOST` - The XNAT server URL
- `XNATR_USER` - Your username
- `XNATR_PASS` - Your password

```{r auth-env}
# Set environment variables (typically done in .Renviron)
Sys.setenv(XNATR_HOST = "https://central.xnat.org")
Sys.setenv(XNATR_USER = "myuser")
Sys.setenv(XNATR_PASS = "mypass")

# Authenticate using environment variables
authenticate_xnat()
```
### Configuration File

For persistent local configuration, create a YAML config file:

```{r auth-config}
# Create template config file
initialize_config()
# This creates ~/.xnatR_config.yml
# Edit the file with your credentials, then:
authenticate_xnat()
```

The config file format:
```yaml
base_url: 'https://your-xnat-server.org'
username: 'your-username'
password: 'your-password'
ssl_verify: true
```

### .netrc File

xnatR also supports the standard `.netrc` file format used by many tools:

```
# ~/.netrc
machine central.xnat.org login myuser password mypass
```

```{r auth-netrc}
# Just provide the URL, credentials come from .netrc
authenticate_xnat(base_url = "https://central.xnat.org")
```

### JSESSION Authentication

For better performance with many requests, use JSESSION-based authentication:

```{r auth-jsession}
authenticate_xnat(
  base_url = "https://central.xnat.org",
  username = "myuser",
  password = "mypass",
  use_jsession = TRUE
)
```

### XNAT 1.8+ Auth Providers

XNAT 1.8+ supports multiple authentication providers (LDAP, OAuth, etc.). Use the `provider/username` format:

```{r auth-provider}
authenticate_xnat(
  base_url = "https://myxnat.org",
  username = "ldap/myuser",
  password = "mypass"
)
```

### Session Management

Check your authentication status:

```{r session}
# Check if authenticated
is_authenticated()

# Get current server URL
xnat_server()

# Get current username
xnat_username()

# Log out
xnat_logout()
```

## Navigating XNAT Data

XNAT organizes data in a hierarchy:

```
Project
└── Subject
    └── Experiment (imaging session)
        └── Scan
            └── Resource (DICOM, NIFTI, etc.)
                └── Files
```

### Listing Projects

```{r list-projects}
# List all projects you have access to
projects <- list_projects()
print(projects)

# Get specific columns only
projects <- list_projects(columns = c("ID", "name", "description"))

# Search for projects by name
mri_projects <- search_projects("MRI")
```

### Listing Subjects

```{r list-subjects}
# List subjects in a project
subjects <- list_subjects(project_id = "MyProject")
print(subjects)

# With pagination for large projects
first_100 <- list_subjects(project_id = "MyProject", limit = 100)
next_100 <- list_subjects(project_id = "MyProject", limit = 100, offset = 100)
```

### Listing Experiments

Experiments represent imaging sessions (MR, PET, CT, etc.):

```{r list-experiments}
experiments <- list_experiments(
  project_id = "MyProject",
  subject_id = "Subject001"
)
print(experiments)
```

### Listing Scans

Scans are individual acquisitions within an experiment:

```{r list-scans}
scans <- list_scans(
  project_id = "MyProject",
  subject_id = "Subject001",
  experiment_id = "Exp001"
)
print(scans)
```

### Listing Resources and Files

Resources are containers for files (e.g., DICOM, NIFTI):

```{r list-resources}
# List resources for a scan
resources <- list_resources(
  project_id = "MyProject",
  subject_id = "Subject001",
  experiment_id = "Exp001",
  scan_id = "1"
)

# List files within a resource
files <- list_files(
  project_id = "MyProject",
  subject_id = "Subject001",
  experiment_id = "Exp001",
  scan_id = "1",
  resource = "DICOM"
)
```

### Derived Data

XNAT stores processed data as assessors and reconstructions:

```{r derived-data}
# List assessors (e.g., FreeSurfer outputs, QC results)
assessors <- list_assessors(
  project_id = "MyProject",
  subject_id = "Subject001",
  experiment_id = "Exp001"
)

# List reconstructions
recons <- list_reconstructions(
  project_id = "MyProject",
  subject_id = "Subject001",
  experiment_id = "Exp001"
)
```

## Downloading Data

### Download Scans

```{r download-scans}
# Download all scans from an experiment
download_files(
  project_id = "MyProject",
  subject_id = "Subject001",
  experiment_id = "Exp001",
  dest_dir = "~/Downloads"
)

# Download specific scans
download_files(
  project_id = "MyProject",
  subject_id = "Subject001",
  experiment_id = "Exp001",
  scan_id = c("1", "2", "3"),
  dest_dir = "~/Downloads"
)

# Download only DICOM files
download_files(
  project_id = "MyProject",
  subject_id = "Subject001",
  experiment_id = "Exp001",
  resource = "DICOM",
  dest_dir = "~/Downloads"
)
```

### Download Subject Data

Download all data for a subject:

```{r download-subject}
download_subject(
  project_id = "MyProject",
  subject_id = "Subject001",
  dest_dir = "~/Downloads"
)
```

### Download Individual Files

```{r download-file}
download_xnat_file(
  url = "/data/projects/MyProject/subjects/S001/experiments/E001/scans/1/resources/DICOM/files/file.dcm",
  dest_file = "file.dcm"
)
```

## Advanced Search

For complex queries across your XNAT data, use the search API.

### Discover Searchable Fields

First, find what data types and fields are available:

```{r search-discovery}
# List all data types
datatypes <- list_data_types()
print(datatypes)

# List queryable fields for MR sessions
fields <- list_queryable_fields("xnat:mrSessionData")
print(fields)
```

### Basic Search

```{r search-basic}
# Search for MR sessions in a project
results <- xnat_search(
  root_type = "xnat:mrSessionData",
  fields = c(
    "xnat:mrSessionData/ID",
    "xnat:mrSessionData/label",
    "xnat:mrSessionData/date"
  ),
  criteria = list(
    list(
      field = "xnat:mrSessionData/project",
      comparison = "EQUALS",
      value = "MyProject"
    )
  )
)
```

### Search Builder (Fluent API)

For complex searches, use the fluent search builder:
```{r search-builder}
results <- xnat_search_builder("xnat:mrSessionData") |>
  search_select(
    "xnat:mrSessionData/ID",
    "xnat:mrSessionData/label",
    "xnat:mrSessionData/date",
    "xnat:mrSessionData/subject_ID"
  ) |>
  search_where("xnat:mrSessionData/project", "EQUALS", "MyProject") |>
  search_where("xnat:mrSessionData/date", "GREATER_THAN", "2023-01-01") |>
  search_execute()

print(results)
```

### Search Criteria

Available comparison operators:

- `EQUALS` - Exact match
- `LIKE` - Pattern match (use % as wildcard)
- `GREATER_THAN` - Greater than
- `LESS_THAN` - Less than
- `GREATER_THAN_EQUAL` - Greater than or equal
- `LESS_THAN_EQUAL` - Less than or equal

```{r search-criteria}
# Create reusable criteria
project_crit <- search_criterion(
  field = "xnat:subjectData/project",
  comparison = "EQUALS",
  value = "MyProject"
)

gender_crit <- search_criterion(
  field = "xnat:subjectData/gender",
  comparison = "EQUALS",
  value = "M"
)

# Use in search
results <- xnat_search(
  root_type = "xnat:subjectData",
  fields = c("xnat:subjectData/ID", "xnat:subjectData/label"),
  criteria = list(project_crit, gender_crit)
)
```

## Alias Tokens

Alias tokens provide temporary credentials that can be revoked without changing your password.

```{r tokens}
# Issue a new token
token <- xnat_token_issue()
print(token$alias)
print(token$estimatedExpirationTime)

# List all your active tokens
tokens <- xnat_token_list()
print(tokens)

# Validate a token
validation <- xnat_token_validate(
  alias = token$alias,
  secret = token$secret
)
print(validation$valid)

# Invalidate a token when done
xnat_token_invalidate(
  alias = token$alias,
  secret = token$secret
)
```

You can use alias tokens for authentication:

```{r token-auth}
# Get a token
token <- xnat_token_issue()

# Use it for authentication in another session
authenticate_xnat(

  base_url = "https://central.xnat.org",
  username = token$alias,
  password = token$secret
)
```

## Working with S3 Classes

All list functions return tibbles with custom S3 classes. These provide nice printing:

```{r s3-classes}
projects <- list_projects()
class(projects)
#> [1] "xnat_projects" "tbl_df" "tbl" "data.frame"

# Custom print shows summary
print(projects)
#> ── XNAT Projects ──────────────────────────────
#> 42 projects
#> # A tibble: 42 × 3
#>    ID          name                    description
#>    <chr>       <chr>                   <chr>
#>  1 CENTRAL_...  Central OASIS ...       ...
#>  ...
```

The objects work like regular tibbles with dplyr and other tidyverse packages:

```{r tidyverse}
library(dplyr)

projects |>
  filter(grepl("MRI", name, ignore.case = TRUE)) |>
  select(ID, name)
```

## Error Handling

xnatR uses cli for informative error messages:

```{r errors}
# Not authenticated
list_projects()
#> Error:
#> ! Not authenticated.
#> ℹ Run `authenticate_xnat()` first.

# Invalid credentials
authenticate_xnat(
  base_url = "https://central.xnat.org",
  username = "wrong",
  password = "wrong"
)
#> Error:
#> ! Authentication failed.
#> ✖ Invalid username or password.
```

## SSL Certificates

For servers with self-signed certificates:

```{r ssl}
authenticate_xnat(
base_url = "https://internal-xnat.myinstitution.org",
  username = "myuser",
  password = "mypass",
  ssl_verify = FALSE
)
```

## Best Practices

1. **Use environment variables** for credentials in scripts
2. **Use alias tokens** for temporary access or sharing
3. **Enable JSESSION** for scripts making many requests
4. **Use pagination** (`limit`, `offset`) for large result sets
5. **Log out** when done, especially with JSESSION

```{r best-practices}
# Example workflow
authenticate_xnat(use_jsession = TRUE)

tryCatch({
  projects <- list_projects()
  # ... do work ...
}, finally = {
  xnat_logout(invalidate_session = TRUE)
})
```

## Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
